---
title: "Module 7 More Dplyr and Tidyr"
author: "Brian P Steves"
date: "`r Sys.Date()`"
output: html_document
---

## dplyr verbs

The dplyr package provides a lot of methods (verbs) for helping to work with data.

### filter()

filter() is equivalent to subset()

```{r}
library(dplyr)
library(Lahman)
# grabbing baseball data from the Lahman package


# select players from the Seattle Mariners in 1990
Batting %>% filter(teamID == "SEA", yearID == 1990) %>% head()

```


### arrange()

arrange() allows you to reorder the data


```{r}

# order data by team then year then by descending number of runs (r) 
Batting %>% arrange(teamID, yearID, desc(R)) %>% head()

```

### select()

select() lets you specify which columns you want to return


```{r}
# select id, year, team and r only
Batting %>% select(playerID, yearID, teamID, R) %>% head()


# select everything but year and team
Batting %>% select(-c(yearID, teamID)) %>% head()


# select id through team
Batting %>% select(playerID:teamID) %>% head()


```

### mutate()
mutate() allows you to add new columns
```{r}
#add a runs per game column and select only the id and runs_per_game column for return
Batting %>% mutate(runs_per_game = R/G) %>% select(playerID, runs_per_game) %>% head()


```
### summarize()
We already showed an example of summarize().  It really needs to be used in conjunction with group_by() to be of much use.  Without group_by() with out it, summarize() returns a single row of results

```{r}
# calculate the mean number of games played per year across all players
Batting %>% summarize(mean_g = mean(G)) 
```


### group_by()

Select() and arrange() aren't affected by group_by() but filter(), mutate(), and summarize() are.

For example can group by team and then summarize

```{r}

Batting %>% group_by(teamID) %>% summarize(mean_g = mean(G)) %>% head()
```


## summary functions

There are a variety of summary functions that can be used with summarize().   These include mean, sd, var, min, max, median, first, last, nth, n, and n_distinct.  The last two n and n_distinct are particularly useful.


n() gives you the count of data rows and doesn't take an arguement
n_distinct() gives you the count of unique values and requires an arguement

```{r}

# number of players by team
Batting %>% group_by(teamID) %>% summarize(count = n()) %>% head()

# can be written as 
Batting %>% group_by(teamID) %>% tally() %>% head()

# also be written as
Batting %>% count(teamID) %>% head()

# number of years each teamID played
Batting %>% group_by(teamID) %>% summarize(year_count = n_distinct(yearID)) %>% head()

```

## window functions

The mutate() verb we mentioned early has a wide range of window functions you can use which are very helpful.

Some shift the data up or down... (lead, lag)

```{r}
measdat <- data.frame(observation = c(1,2,3,4,5,6), size = c(10, 13, 11, 17, 10, 9) )

#shifting up
measdat %>% mutate(next_size = lead(size))

#shifting down
measdat %>% mutate(last_size = lag(size))

```


Some are used to rank data... (dense_rank, min_rank, percent_rank, row_number, ntile, between, ntile)

```{r}
# rank the size measurements
measdat %>% mutate(dense_rank(size))

```


Some do cumulative values .. (cumall, cumany, cummean, cumsum, cummax, cumprod)

```{r}

# running mean of size measurements
measdat %>% mutate(cummean(size))

```


## join
If you need to combine data from two tables based on a common field you can use the various join functions. The basic format is like the following...

If you don't specify a "by" then dplyr will match based on common column names

```{r}
# two small data frames for the following examples
x <- data.frame(id=c(1,2,3,4,5,6), name=c("Fred","Bob","Jim","Ted","Jane","Sue"))
x

y <- data.frame(id=c(1,2,5,6,7,8), score=c(78,84,92,83,45,93))
y

```


### inner_join()
inner_join() returns all columns from where the two data frames have matching records.

```{r}

inner_join(x,y, by ="id")

```

### left_join()
left_join() returns all columns from the two data frames.  But it returns all the records from the first data frame (the left one) and only those from the second data frame where there is a match

```{r}

left_join(x,y, by ="id")

```

### semi_join()
semi_join() returns all rows from the first data frame where there are matching values in the second data frame and it only keeps the columns from the first data frame.

```{r}

semi_join(x,y, by ="id")

```
### anti_join()
anti_join() returns all the rows from the first data frame where there are no matches in the second data frame and it only keeps the columns in the first data frame.

```{r}

anti_join(x,y, by ="id")

```








