---
title: 'Module 4:  Filtering and Subsetting Data'
author: "Brian P Steves"
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: yes
---


We briefly discussed selecting data within an object in Module 2.  Here we will discuss more advanced methods to subsetting data. Because most of the data we will be using in R will be in the form of data frames we will concentrate on data frame methods for filtering and subsetting.


##  Determining the structure of an object

In order to understand how to select portions of a data object, you will need to know details about the structure of that object first.  The str() function accomplishes that.

```{r}
# Create a series of equally sized vectors
sites<-c("Site A", "Site B")
dates<-as.Date(c("12Mar2013","04Apr2013"),"%d%b%Y")
lat=c(45, 47)
lon=c(-124, -125)
treatment=c(T,F)

# Combine the vectors into a data frame
dat<-data.frame(sites, dates, lat, lon, treatment)

# Use str() on some of the vectors and the data frame
str(sites)
str(lat)
str(dat)

# The name() function is similar but only provides the names of the various data frame columns
names(dat)


# You can also use name() to reassign the names of the data frame columns

names(dat)<-c("SiteName", "VisitDate", "Latitude", "Longitude", "TreatmentApplied")

str(dat)

```


## Head and Tail

Large data frames take a long time to print out to the screen.  Because of this, sometimes you just want to view the first portion (the head) or the last portion (the tail) of a data frame.  To get a sense of what the data looks like.   

```{r}

data(iris)

# How many rows of data using nrow()
nrow(iris)

# too much to show on screen


# The head (beginning) of the data
head(iris)


# The tail (end) of the data
tail(iris)

# the defualt is 6 rows of data, but you can change this by specifing an n value

head(iris, n=20)

```

##  Selecting Variables (data frame columns)

### Selecting Single columns using Base R

```{r}

# By index
dat[1]

# By Name
dat["SiteName"]

# Using the $ operator

dat$SiteName


```

### By Name

``` {r}

dat[c("SiteName","TreatmentApplied")]

# Alternatively you can use matrix R,C notation

dat[,c("SiteName","TreatmentApplied")]

```

### By column number like a matrix

```{r}

dat[,c(1,5)]

```

### By Logical Assignment

```{r}
dat[c(T,F,F,F,T)]

```

##  Droping Variables

### Excluding Variables

```{r}

# we can use negative indexes

dat[c(-1,-4)]

# or we can use logical assignment

dropcols<-names(dat) %in% c("Latitude","Longitude")
dropcols

dropcols2<-c("Latitude","Longitude")

# this is the similar to a logical assignment for selecting columns but instead a "True" is a column we want to drop 
dat[c(F,F,T,T,F)]

# We use the ! to denote a logical negation or "not"
# Here we specify that we want to select "not the dropcols"
dat[!dropcols]

```


### Dropping Variables permentantly

```{r}

# One option is reassignment

dat<-dat[!dropcols]

# rebuild the data frame
dat<-data.frame(sites, dates, lat, lon, treatment)
names(dat)<-c("SiteName", "VisitDate", "Latitude", "Longitude", "TreatmentApplied")

# the other option is to set some columns to null

dat$TreatmentApplied <-NULL
dat
# But you can't do this with more than one column at a time

dat[c("Latitude","Longitude")]<-NULL

# However you can chain these

dat$Latitude<-dat$Longitude<-NULL
dat


```

## selecting columns using Tidyverse's 'dplyr' package.  

```{r}
library(dplyr)
dat<-data.frame(sites, dates, lat, lon, treatment)
names(dat)<-c("SiteName", "VisitDate", "Latitude", "Longitude", "TreatmentApplied")

select(dat, SiteName, Latitude)

```

Note that the first input to the select function is the name of the data frame.

The dplyr package uses a new type of notation called piping using "%>%" between expressions. This eliminates the need to nest functions within each other and streamlines a lot of coding.

```{r}

dat %>% select(Latitude, Longitude)
```
Note that we start with the data frame and then pipe in the 'select' function. In doing so we don't need to include the data frame name as the first input of the the select function. The result of this bit of code can then be piped into more dplyr expressions (more on that later)


The dplyr select() function will accept a wide range of helper functions that help select columns based on criteria.

```{r}

head(iris)

iris %>% select(contains("Sepal")) %>% head()

iris %>% select(contains("Width")) %>% head()

# all columns except "Species"
iris %>% select(-Species) %>% head()

# select Species and then the two middle measurement columns
iris %>% select(Species, Sepal.Width:Petal.Length) %>% head()

```



##  Selecting Observations (data frame rows) firs using Base R

``` {r}
# loading some real data with lots of records
library(readr)
mtcars <- read_csv(readr_example("mtcars.csv"))

str(mtcars)

```

### By index

```{r}

# select first two records
mtcars[1:2,]

# selecting second and fourth records
mtcars[c(2,4),]

# selecting not the third and fifth record
mtcars[c(-3,-5),]

```

### By value

``` {r}

# Selecting cars with 4 cylinders
mtcars[mtcars$cyl == 4,]

# Using & to combine conditions
mtcars[mtcars$cyl == 4 & mtcars$mpg > 30,]

```

##  Selecting Observations and Variables

``` {r}
# mpg and hp for 4 cylinder cars 

mtcars[mtcars$cyl == 4,c("mpg","hp")]

# Number of gears for 4 cylinder cars
mtcars[mtcars$cyl == 4,]$gear
```

##  Subset Function

The subset() function is a shortcut to selecting observations and variables

``` {r}
# this isn't any shorter than the previous method of finding the cars based on cyl, hp, mpg.

subset(mtcars, cyl == 4, select=c(mpg, hp))

# we can also drop columns using the select option

subset(mtcars, cyl==4, select=c(-mpg, -hp))

# Subset makes it easier to combine multiple conditions

subset(mtcars, cyl==4 & mpg > 30 & hp > 100)

# You can even do "OR" conditions using the "|" operator
# here we are selecting where cyl  = 4 or mpg < 15

subset(mtcars, cyl ==4 | mpg < 15)




```
 Remember you can combine a series of logical operators ( "|" and "&")
 it helps if you know some logic to help simply your expressions
 for example...
 
 !(a & b) is the same as !a | !b
 !(a | b) is the same as !a & !b 

```{r}

 subset(mtcars, !(cyl  == 4 & gear == 4))
 
 # is the same as 

 subset(mtcars, !cyl == 4 | !gear == 4 )

 # and likewise

 subset(mtcars, !(cyl == 4 | gear == 4))

 # is the same as 

 subset(mtcars, !cyl == 4 & !gear == 4)



``` 

You can also use the %in% operator to match multiple values at once.

```{r}

hp_of_interest<-c(110, 180, 245)
subset(mtcars, hp %in% hp_of_interest)
 
# is the same as 
subset(mtcars, hp == 110 | hp == 180 | hp == 245)


```

## subsetting data using tidyverse's dplyr

Use commas to separate "and" statements.
Use | to separate "or" statements.

```{r}
mtcars %>% filter(hp == 110)

mtcars %>% filter(hp < 110,cyl == 4, wt>3)

mtcars %>% filter(hp > 100 | cyl == 4)

mtcars %>% filter(hp %in% hp_of_interest)
```

## combining dplyr filter and select

```{r}
mtcars %>% filter(cyl == 4) %>% select(cyl, mpg, hp)
```




##  Random samples in Base R
Sometimes you will want a random subset of your data.  To do this, use the sample() function without replacement.

```{r}

# first determine how many rows of data you have
rowcount<-nrow(mtcars)

# select two random rows
randomrows<-sample(rowcount, 2)

# use the "randomrows" to select those rows from the data frame
mtcars[randomrows,]


``` 

## Select random rows in dplyr using slice_sample() function

We'll randomly select two rows and do it twice so show that they are random

```{r}
mtcars %>% slice_sample(n=2)
mtcars %>% slice_sample(n=2)
```
Can also slice a proportion of the data.  Here 10% of 32 records returns 3 random records
```{r}
mtcars %>% slice_sample(prop=0.1)
```



## Head, Tail, Glimpse with dplyr

head and tail can be passed into the pipe using %>%

```{r}

iris %>% head()

iris %>% tail()

iris %>% head(n=20)


```

dplyr also has the "glimpse()" function that gives an alternative view of the data

```{r}
iris %>% glimpse()

```


##  Assigning and Subsetting

Sometimes you want to select data so that you can reassign values in the data frame. This is useful when you are cleaning up your data.

```{r}

# for example, sometimes you want to replace NA values with 0's

data3<-data.frame(site=c(1,1,1,2,2,2),count=c(2,4,3,NA,5,NA))
data3

# to select rows where count is NA you can use the is.na() function

data3[is.na(data3$count),]

# to set count to zero where count was NA

data3[is.na(data3$count),]$count<-0

data3

```

This is much easier with dplyr using mutate() and replace()

The usage is replace(variable, condition, replacement)

```{r}
data3<-data.frame(site=c(1,1,1,2,2,2),count=c(2,4,3,NA,5,NA))

data3 %>% mutate(count = replace(count, is.na(count), 0))
data3 %>% mutate(count = replace(count, site==2, 100))
```

Mutate can also be used to create new variables
```{r}

iris %>% mutate(Sepal.Area = Sepal.Length * Sepal.Width)

```



##  Ordering

Ordering data in base R is annoying. I'm not even going to bother showing it to you. But it's pretty simple with dplyr's arrange() function.

```{r}

# ascending
mtcars %>% arrange(mpg)  

#descending
mtcars %>% arrange(desc(mpg))


```


## Bonus Material

Here's a cheat sheet for wrangling data with dplyr.  

https://rstudio.github.io/cheatsheets/data-transformation.pdf

We still have much to cover with dplyr as well as the related package called tidyr.

## Homework

Using the data set you found, write  R code to create 3 interesting subsets of data that you might want to use for later analysis.  Use which ever methods make the most sense to you but be sure to assign these subsets to new variables with meaningful names. Remember to add code to import this data first.




